**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**

### 2023-02-07 data_cleaning/Data_cleaning_final.qmd
created file with details of data cleaning

---
title: "Fields COVID-19 dataset exploration"
author: "Ariel Mundo"
format: pdf
toc: true
toc-depth: 2
editor: source
urlcolor: blue
#bibliography: refs.bib
---

@@

This dataset is about vaccination and other COVID measures in different cities in Ontario. The covariates available are:

-   Age (only above 16, if age above 75 then it appears as 98)
-   Age-group (generated from age)
-   Employment status
-   Remote work within the last month
-   If person receives paid sick leave
-   Number of people in household
-   Number of people from household that attend school
-   Chronic illnesses within the household (*I will not analyze this one as the question is too open and asks about age and disease at the same time*)
-   Race
-   Three first digits of postal code
-   Day, month and year the survey was accessed

There are many responses from the survey on the dataset, but there are some that I would be interesed in analyzing:

@@

**Note: As of Dec 30, 2022 I no longer intend to analyze this last question as after cleaning the dataset it has a missing rate of about 76% and therefore it has too few observations compared to the first two questions**.

@@

This document focuses on some preliminary analyses of the data I undertook to assess its representativeness. I performed five major tasks for this preliminary analysis:

-   Choosing covariates to assess their missigness rates
-   Cleaning the data in order to keep all the covariates with a 0% missigness rate
-   Race and ethnicity representativeness of the clean data
-   Trends in the survey answers by geographical location
-   Trends in survey answers by group age

Overall the goal of this analysis was to see what adjustments would need to be made to formally analyze the data.

@@

@@

#set the time and date fields in the format that is required by lubridate

data$time_date_code<-ymd(data$time_date_code)
```

```{r, rename-columns,echo=FALSE,warning=FALSE}

#get a new dataframe and extract variables of interest and make them factors

#get the variables of interest, convert to factors those required
d1<-data%>%select(
  time_date_code,
  city,
  age,
  age_group,
  q02_first_dose,
  q02a_second_dose,
  q11_employed,
  q13_hh_size,
  q09_hh_income,
  q02b_vaccine,
  q16_race,
  q11a_work_remotely,
  q11b_sick_leave,
  q14_hh_attending_school,
  q24a_postal_first,
  q24b_postal_second,
  q24c_postal_third)%>%
  rename(date=time_date_code,
         first_dose=q02_first_dose,
         second_dose=q02a_second_dose,
         get_vaccine=q02b_vaccine,
         h_size=q13_hh_size,
         income=q09_hh_income,
         employed=q11_employed,
         race=q16_race,
         remote_work=q11a_work_remotely,
         sick_leave=q11b_sick_leave,
         school=q14_hh_attending_school,
         pc_1=q24a_postal_first,
         pc_2=q24b_postal_second,
         pc_3=q24c_postal_third
         )%>%
  mutate(across(c("age_group",
            "city",
            "employed",
            "h_size",
            "income",
            "race",
            "remote_work",
            "sick_leave",
            "school"),as.factor))

@@

The next step is to see from the covariates what are the ones with the highest number of missing observations, in order to decide what can we include in the model.

@@

However, there are a couple of things we want to consider: The answer to "sick_leave" was answered only by those that reported to be employed (the survey design made this response conditional). Therefore, those unemployed would be excluded in an analysis that considers this variable.

At the very least, it would be interesting to analyze how the reported status of vaccination changed over geographical location and time by:

@@

From these covariates, the one with the highest missing rate of observations is the employment status. I next cleaned the data in order to have complete observations about employment, and to see how the missing rates looked for the other covariates.

@@

a<-d1%>%
  mutate(income_ord = fct_relevel(income,
                                   "under_15000",
                                   "15000_24999",
                                   "25000_39999",
                                   "40000_59999",
                                   "60000_89999",
                                   "90000_109999",
                                  "over_110000"
                                   ))%>%
 #drop_na(age,income,h_size)%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~income_ord)+
  ylim(c(0,1200))+
  theme(strip.text.x = element_text(size = 8))+
  theme_classic()+
  thm1

b<-d1%>%
  mutate(h_size_ord = fct_relevel(h_size,
                                   "1",
                                   "2",
                                   "3",
                                   "4",
                                   "5",
                                   "6",
                                   "7",
                                   "8",
                                   "9",
                                   "10_or_more"
                                   ))%>%
  #drop_na(age,income,h_size_ord)%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~h_size_ord)+
  ylim(c(0,200))+
  theme(strip.text.x = element_text(size = 8))+
  theme_classic()+
  thm1

@@

d<-clean_data%>%
   mutate(h_size_ord = fct_relevel(h_size,
                                   "1",
                                   "2",
                                   "3",
                                   "4",
                                   "5",
                                   "6",
                                   "7",
                                   "8",
                                   "9",
                                   "10_or_more"
                                   ))%>%
  #drop_na(age,income,h_size_ord)%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~h_size_ord)+
  ylim(c(0,200))+
  theme_classic()+
  theme(strip.text.x = element_text(size = 8))+
  thm1

@@

e<-clean_data%>%
  #drop_na(age,income,h_size_ord)%>%
  ggplot(aes(x=age,fill=employed))+
  geom_histogram(stat="count",binwidth=10)+
  #ylim(c(0,200))+
facet_wrap(~income)+
  theme_classic()+
  ggtitle("Clean dataset")+
  #theme(strip.text.x = element_text(size = 8))+
  thm1

f<-d1%>%
  #drop_na(age,income,h_size_ord)%>%
  ggplot(aes(x=age,fill=employed))+
  geom_histogram(stat="count",binwidth=10)+
  #ylim(c(0,200))+
facet_wrap(~income)+
  theme_classic()+
  ylim(c(0,100))+
  ggtitle("Original dataset")+
  #theme(strip.text.x = element_text(size = 8))+
  thm1

@@

There seem to be outliers where people \<25 y of age report having an income \>110k while living in a household of 1. Explore the reported household composition of those \<25 with income\>110k.

@@

This table shows something interesting due to the nature of the dataset: the missing observations across variables are different, which is why now, although in the original dataset there were more observations of "race" than "employment", when we select complete observations for the latter we lose some observations for race. The second step of cleaning would be to remove the missing observations from race as well, and see what the proportions of covariates and responses are when compared to the census data.

@@

It can be seen that after removing all missing observations in the covariates, the only response that still has missing observations is the answers for the second dose of the vaccine, with 27% of missing observations.

@@

Once the data was clean, I explored how the race and ethnicity information from the dataset compared to the data from the Census.

@@

Because the categories for race/ethnicity in the survey did not match the categories from the Canadian Census, we used a combination of sources to obtain the corresponding percentages to match the survey categories. Sources are indicated below:

@@

| Ethnicity/Race | percentage |
|----------------|:-----------|
| Arab           | 1.6%       |
| Black          | 4.7%       |
| East Asian/Pacific Islander     | 6.6%       |
| Indigenous     | 2.8%       |
| Latin American | 1.5%       |
| Mixed          | 1.0%       |
| Other          | 5.2%       |
| South Asian    | 8.7%       |
| White Caucasian| 67.8%      |
| **Total**      | 99.9%      |
: Reference Data from Race/Ethnicity

@@

Another important aspect to consider is how the trends in the responses look over time by age group. This is because from the census data, the age groups for the province are as follows:

@@

However, when plotted the proportion of answers at each time point among groups, the distribution looks rather different.

@@

It can be seen from the graph that the proportion of answers by group age do not follow the overall distribution from the province.

@@

The question from the survey was "What is your household annual income?". To compare the results of this answer, we used the "Household total income groups in 2015 for private households" from the Census data (available in the Census Data for Ontario website [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Income&TABID=1&type=1)).

@@

The brackets for income in the census data are different than the brackets used in the survey. The census does CAD 4,999 brackets (e.g., CAD 5,000- CAD 9,9999) up to CAD 49,999, followed by CAD 9,999 brackets up to CAD 99,999. After that, the brackets increase to CAD 24,999, and therefore, one cannot obtain percentages for the 90,000-109,999 and \>110,000 brackets from the survey.

Therefore, an additional category for income will be created in the dataset so it matches the information from the Census.

@@

Again, there are differences here in the patterns of response. The \<15,000 bracket accounts for about 20% of the responses, a much higher rate than the percentage they represent from the census, and there is variation within the other brackets as well.

## Geographical location

Next, we are interested in analyzing the geographical distribution of the answers the participants within the geographical areas defined by the Association of Municipalities from Ontario.

**Update: the populations by census division/city were found on Jan 8, 2023 (from https://www12.statcan.gc.ca/census-recensement/2021/dp-pd/prof/index.cfm?Lang=E). The correction will be done using these numbers instead. The numbers can be found in `census_population_by_division.csv` in `data` directory.**

@@

1) Add to the dataset the data for the municipalities so each city has the corresponding region it belongs to.
1) Determine percentage of answers that correspond to Toronto. This is important because although the Census regions do not match the exact geographical boundaries of the municipalities and thus a direct correction for this cannot be easily obtained, Toronto is the largest urban area in the province and it is important to identify if corrections are needed with regard to the proportions from this place in the survey.

### Joining the municipalities data

@@

Need to explore which entries were left without a geographical region:

@@

There are 2744 observations that did not get a geographical region, and they correspond to 187 unique cities. These will be exported to a csv file in order to manually write their geographical areas following the Association of Municipalities of Ontario divisions, and using Wikipedia to check the status of each municipality.

@@

The code chunk above was ran once to get the names of the municipalities. After manually entering the geographic areas, the file was saved as `missing_municipalities_updated.csv`, and this will be the file to be used for the next steps.

Note: there is one city "Kinburn", but there are two communities with such name, one in Huron County and one in Carelton County, for the time being Huron County is assigned, will check later if is better to remove it.

Also one occurence of "Sydenham", which can be a ward in Kingston or a community in Frotenac. Will assign Frotenac for the time being.

@@

After manually assigning the geographical regions to the municipalities that were missing (which can be found in `missing_municipalities.csv` in the `data` directory), these will be merged as the dataset `geographic_areas.csv` (also in `data`), which contains the titles for each region (e.g., "County", "Region").

@@

## Proportion of answers within Toronto

**This section needs to be modified as the correction has been done by geographical area on Jan 8, 2023**

The final step is of the geographical analysis is to determine the proportion of answers from Toronto.

```{r,plot-location}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
#| fig-cap: Percentage of survey answers by city

clean_data<-clean_data%>%
  mutate(location=as.character(case_when(city=="Toronto"~"Toronto",
                            city!="Toronto"~ "All other cities"
                          )))%>%
  mutate(location=as.factor(location))

ggplot(clean_data %>% count(location) %>%    # then count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(location, n, fill=location)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm1

The proportion of responses from the city of Toronto is 31.3% of the total in the clean dataset. From the (Census data)[https://www.toronto.ca/wp-content/uploads/2022/02/92e3-City-Planning-2021-Census-Backgrounder-Population-Dwellings-Backgrounder.pdf], the city of Toronto as of 2021 has a population of 2,794,356, corresponding to 19.6% of the total population of the Province.

# Health Regions approach

Following on a suggestion by BN on Jan 09, the idea is to use broader regions for geographical analysis. The Health Regions of Ontario will be used. However, these Health Regions do not match the census divisions. I have not been able to find a dataset from Health Ontario that lists each municipality and its corresponding Health Region. I will use a multi-stage approach to incorporate the information into the dataset:

- Get dataset from Paul Allen regarding long-term care homes in Ontario (https://paulallen.ca/consolidated-dataset-of-ltc-homes-in-ontario/). This dataset has communities and the Local Health Integration Network (LHINs) where long-term care homes were located. I will keep the community and LHIN information.

- Add column for Health Region based on LHIN: The Health Regions substituted LHINs, a list of their correspondence can be found here: https://www.ontariohealth.ca/about-us/our-people. I will add this information.

- Merge the dataset and check if there are missing municipalities. Manually add missing municipalities.

## Steps

The dataset from Paul Allen was downloaded and saved as `Consolidated_LTC_dataset.csv` in the `data` directory. One thing to note is that there is a missing observation (coded as "Not provided") for one of the entries (city of Napanee, located in the Lennox and Addington County). Under the LHIN divisions, Napanee was in the South East LHIN. Also, there is one entry that says "244 Main Street East" as the community entry but it should be "Stayner" (the address provided belongs to Stayner.) The information is fixed before adding the Health Region.

@@

Now, combine datasets and write csv file for those cities that were not assigned a Health Region.

@@

To complete the missing LHIN information, the correspoding LHINs were obtained from the LHIN websites:

https://www.southeasthealthline.ca/aboutourregion.aspx?region=RuralFrontenacLennoxAddington

https://www.nsmhealthline.ca/aboutourregion.aspx

https://www.champlainhealthline.ca/aboutourregion.aspx?region=StormontDundasGlen

https://www.wwhealthline.ca/aboutourregion.aspx?region=KitchenerWaterlooWellesleyWilmotWoolwich

Some cities did not belong entirely to a LHIN. For example, Etobicoke was divided between the Central, Central East, and Toronto Central LHINs. We chose in these cases the LHIN that covered the larger geographical region of each city (Toronto Central LHIN in the case of Etobicoke). We now assign LHINs to the dataset and Health Regions. The dataset with the information of the missing LHINs is called `missing_health_regions_updated.csv`

@@

At this point, the data has been cleaned and prepared for formal analysis.

The clean dataset will be saved as a *.csv file.

@@

write.csv(clean_data,here("data","clean_dataset.csv"),row.names = FALSE)
```

>>>>

---
title: "**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**"
subtitle: Data Cleaning
author: "Ariel Mundo"
format: pdf
toc: true
toc-depth: 2
editor: source
urlcolor: blue
#bibliography: refs.bib
---

@@

The original Fields Covid-19 survey contained information about Covid-19 vaccination status and other COVID measures in different cities in Ontario. Information provided by respondents included:

-   Age (only above 16, if age above 75 then it appears as 98)
-   Age-group (generated from age)
-   Employment status
-   Remote work within the last month
-   If person receives paid sick leave
-   Number of people in household
-   Number of people from household that attend school
-   Chronic illnesses within the household same time
-   Race
-   Three first digits of postal code
-   Day, month and year the survey was accessed

Respondents provided multiple answers regarding vaccination:

@@

@@

This document focuses on exploratory analyses, covariate selection, and outlier identification and removal. The goal of the exploratory analyses was to clean the dataset for formal analyses.

@@

## Data Loading

The first task was to load the raw data, extract the majority of the responses from the survey that could be used for regression, and if they were categorical, make them factors to do the exploratory analysis.

@@

#get the variables of interest, convert to factors when required
d1<-data%>%select(
  time_date_code,
  city,
  age,
  age_group,
  q02_first_dose,
  q02a_second_dose,
  q11_employed,
  q13_hh_size,
  q09_hh_income,
  q02b_vaccine,
  q16_race,
  q11a_work_remotely,
  q11b_sick_leave,
  q14_hh_attending_school,
  q24a_postal_first,
  q24b_postal_second,
  q24c_postal_third)%>%
  rename(date=time_date_code,
         first_dose=q02_first_dose,
         second_dose=q02a_second_dose,
         get_vaccine=q02b_vaccine,
         h_size=q13_hh_size,
         income=q09_hh_income,
         employed=q11_employed,
         race=q16_race,
         remote_work=q11a_work_remotely,
         sick_leave=q11b_sick_leave,
         school=q14_hh_attending_school,
         pc_1=q24a_postal_first,
         pc_2=q24b_postal_second,
         pc_3=q24c_postal_third
         )%>%
  mutate(across(c("age_group",
            "city",
            "employed",
            "h_size",
            "income",
            "race",
            "remote_work",
            "sick_leave",
            "school"),as.factor))

@@

The next step consisted in identify the missing rates of the covariates, and determinewhich of those could be included in the model. The following code chunk creates a table with the number of missing observations and percentages.

@@

However, per the dictionary in the original dataset, "sick_leave" was answered only by those that reported to be employed (the survey design made this response conditional). Therefore, those unemployed would be excluded in an analysis that considers this variable.

Therefore, we decided to select the following covariates from the original dataset:

@@

From these covariates, the one with the highest missing rate of observations is the employment status. Next, we cleaned the data in order to have complete observations about employment, and to see how the missing rates looked for the other covariates. The following code chunk creates histograms for the different covariates and a table with missing rates in the the `clean_data` data frame which was created from the raw data.

@@

a<-d1%>%
  mutate(income_ord = fct_relevel(income,
                                   "under_15000",
                                   "15000_24999",
                                   "25000_39999",
                                   "40000_59999",
                                   "60000_89999",
                                   "90000_109999",
                                  "over_110000"
                                   ))%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~income_ord)+
  ylim(c(0,1200))+
  theme(strip.text.x = element_text(size = 8))+
  theme_classic()+
  thm1

b<-d1%>%
  mutate(h_size_ord = fct_relevel(h_size,
                                   "1",
                                   "2",
                                   "3",
                                   "4",
                                   "5",
                                   "6",
                                   "7",
                                   "8",
                                   "9",
                                   "10_or_more"
                                   ))%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~h_size_ord)+
  ylim(c(0,200))+
  theme(strip.text.x = element_text(size = 8))+
  theme_classic()+
  thm1

@@

d<-clean_data%>%
   mutate(h_size_ord = fct_relevel(h_size,
                                   "1",
                                   "2",
                                   "3",
                                   "4",
                                   "5",
                                   "6",
                                   "7",
                                   "8",
                                   "9",
                                   "10_or_more"
                                   ))%>%
  ggplot(aes(x=age))+
  geom_histogram(binwidth=10)+
  facet_wrap(~h_size_ord)+
  ylim(c(0,200))+
  theme_classic()+
  theme(strip.text.x = element_text(size = 8))+
  thm1

@@

e<-clean_data%>%
  ggplot(aes(x=age,fill=employed))+
  geom_histogram(stat="count",binwidth=10)+
facet_wrap(~income)+
  theme_classic()+
  ggtitle("Clean dataset")+
  thm1

f<-d1%>%
  ggplot(aes(x=age,fill=employed))+
  geom_histogram(stat="count",binwidth=10)+
  #ylim(c(0,200))+
facet_wrap(~income)+
  theme_classic()+
  ylim(c(0,100))+
  ggtitle("Original dataset")+
  thm1

@@

At this stage in the analysis, we identified that outliers were present in the data. Specifically, individuals that were under 25 years of age and that reported having an income \>110k while living in a household of 1. We therefore, created some diagnostic plots to explore the reported household composition of those \<25 with income\>110k, and we removed the outliers from the dataset.

@@

The next code chunk creates another table with covariate missing rates in the clean dataset. It can be seen that at this stage the answer about the first dose of the vaccine has no missing observations, the answer about having received the second dose of the vaccine has 27% of missing observations, and the answer about if people would get a vaccine has 76% of missing observations.

@@

@@

Next, we explored how the race and ethnicity information from the dataset compared to the data from the Census. The following chunk creates a summary table for the Race variable.

@@

It is important to mention that the categories for race/ethnicity provided in the survey did not match the categories used in the Census. Therefore, we used a combination of sources to obtain the information presented in @tbl-races, where the categories used could be matched to the Census Information. The data sources for the table were:

@@

| Ethnicity/Race | percentage |
|----------------|:-----------|
| Arab           | 1.6%       |
| Black          | 4.7%       |
| East Asian/Pacific Islander     | 6.6%       |
| Indigenous     | 2.8%       |
| Latin American | 1.5%       |
| Mixed          | 1.0%       |
| Other          | 5.2%       |
| South Asian    | 8.7%       |
| White Caucasian| 67.8%      |
| **Total**      | 99.9%      |
: Reference Data from Race/Ethnicity {#tbl-races}

@@

According to the Census, the distribution of the different the age groups for the province of Ontario are as follows:

@@

To compare the survey data to the Census, the next chunk creates a barplot of the age groups in the survey. Here, it can be noticed that the age-group distribution from the survey data is different from the Census data.

@@

@@

Survey respondents answered the question "What is your household annual income?". To compare the distribution of responses to this question from the survey and the Census data, we used the "Household total income groups in 2015 for private households" from the Census data (available in the Census Data for Ontario website [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Income&TABID=1&type=1)).

@@

One difference that is noticeable is that the brackets for income in the census data are different than the brackets used in the survey. The census does CAD 4,999 brackets (e.g., CAD 5,000- CAD 9,9999) up to CAD 49,999, followed by CAD 9,999 brackets up to CAD 99,999. After that, the brackets increase to CAD 24,999, and therefore, it is not possible to obtain percentages for the 90,000-109,999 and \>110,000 brackets from the survey.

Therefore, in the following code chunk, we created an additional category for income in the dataset that matched the information from the Census, and barplots to visualize the proportion of each income bracket in the dataset.

@@

We identified differences again between the survey data and the Census data. For example, the \<15,000 bracket accounts for about 25% of the responses, a much higher rate than the proportion in the Census. The other brackets were different as well from the Census data.

## Geographical Information

Because each of the respondents of the survey was assigned a geographical location (city), we were interested in accounting for geographical location in our analysis. We used a multi-step process to assign geographical information to the entries in the dataset.

@@

1) Assign to each entry in the dataset municipalities the geographical region it belongs to using municipality and geographical region information.

2) Assign a Health Region to each survey entry using the geographical and Local Integrated Health Network (LHIN) information.

The details of each step are outlined below.

### Municipality Data

@@

The following chunk identified which entries were left without a geographical region:

@@

This analysis identified that 2744 entries did not get a geographical region. These 2744 entries corresponded to 187 unique cities. These cities without a region were exported to a csv file in order to manually write their geographical areas following the Association of Municipalities of Ontario divisions, using Wikipedia to check the status of each municipality. The following chunk created the csv file. The code is commented now as it was run once.

@@

After searching and manually entering the geographic area for each city, the file was saved as `missing_municipalities_updated.csv`, and this file was used for the next steps.

Note: there is one city "Kinburn", but there are two communities with such name, one in Huron County and one in Carelton County, we assigned it to Huron County. In the case of "Sydenham", which can be a ward in Kingston or a community in Frotenac, we assigned it to Frotenac.

@@

After manually assigning the geographical regions to the municipalities that were missing (which can be found in `missing_municipalities.csv` in the `data` directory), these were merged as the dataset `geographic_areas.csv` (also in `data`), which contains the titles for each region (e.g., "County", "Region").

@@

### Health Regions

We sought to geographically analyze the information in the survey using the The Health Regions of Ontario. However, these Health Regions do not match the divisions from the census, and there is no publicly available dataset from Health Ontario that lists each municipality and its corresponding Health Region. We used therefore a multi-stage approach to incorporate the information into the dataset:

- First, we used the dataset from Paul Allen regarding long-term care homes in Ontario (https://paulallen.ca/consolidated-dataset-of-ltc-homes-in-ontario/) to obtain information about communities and the Local Health Integration Network (LHINs) where long-term care homes were located.

- Second, using the LHIN information, we added the Health Region each entry corresponded to using the information on LHIN and Health Region correspondence, which can be found here: https://www.ontariohealth.ca/about-us/our-people.

- Third, after merging the dataset, we manually added LHINs to those municipalities that did not have an entry at this stage.

The dataset from Paul Allen was downloaded and saved as `Consolidated_LTC_dataset.csv` in the `data` directory. One thing to note is that there was a missing observation (coded as "Not provided") for one of the entries (city of Napanee, located in the Lennox and Addington County). Under the LHIN divisions, Napanee was in the South East LHIN. Also, there is one entry that says "244 Main Street East" as the community entry but it should be "Stayner" (the address provided belongs to Stayner). The information was fixed before adding the Health Region.

@@

Next, we combined the datasets and wrote csv file with those cities that were not assigned a Health Region. The next code chunk does these steps (note that the line for writing the csv file has ben commented as it was run only once).

@@

To obtain the missing LHINswe, obtained information from the LHIN websites, which listed all the municipalities within each LHIN. The websites were:

- South East [link](https://www.southeasthealthline.ca/aboutourregion.aspx?region=RuralFrontenacLennoxAddington)

- North Simcoe Muskoka [link](https://www.nsmhealthline.ca/aboutourregion.aspx)

- Champlain [link](https://www.champlainhealthline.ca/aboutourregion.aspx?region=StormontDundasGlen)

- Waterloo Wellington [link](https://www.wwhealthline.ca/aboutourregion.aspx?region=KitchenerWaterlooWellesleyWilmotWoolwich)

- North West [link](https://www.northwesthealthline.ca/aboutourregion.aspx)

- North East [link](https://www.northeasthealthline.ca/aboutourregion.aspx)

- Erie St. Clair [link](https://www.eriestclairhealthline.ca/aboutourregion.aspx)

- South West [link](https://www.southwesthealthline.ca/aboutourregion.aspx)

- Hamilton Niagara Haldimand Brant [link](https://www.hnhbhealthline.ca/aboutourregion.aspx)

- Central West [link](https://www.centralwesthealthline.ca/aboutourregion.aspx)

- Central East [link](https://www.centraleasthealthline.ca/aboutourregion.aspx)

- Mississauga Halton [link](https://www.mississaugahaltonhealthline.ca/aboutourregion.aspx?region=HaltonHills)

- Toronto Central [link](https://www.torontocentralhealthline.ca/aboutourregion.aspx)

Some cities did not belong entirely to a LHIN. For example, Etobicoke was divided between the Central, Central East, and Toronto Central LHINs. We chose in these case the LHIN that covered the larger geographical region of each city (Toronto Central LHIN in the case of Etobicoke). We next assigned LHINs to the entries were they were missing using the information from the webistes, and created a csv with the updated information (the file is called `missing_health_regions_updated.csv`). The next code chunk loads this file, assigns LHINs and creates a new column in the dataset for Health Region.

@@

At this point, the data was ready for formal analysis.

The clean dataset and completed dataset was saved as a *.csv file called `clean_dataset`.

@@

#write.csv(clean_data,here("data","clean_dataset.csv"),row.names = FALSE)
```

### 2023-02-07 manuscript/02-methods.qmd
created file with details of data cleaning

| Variable             | Values                                                                                                                          |
|:------------------|-----------------------------------------------------|
| Age group            | 15-24,25-34,35-44,45-54,55-64, 65+                                                                                              |
| Income bracket (CAD) | \<15,000, 15,000-24,999, 25,000-39,999, 40,000-59,999, 60,000-89,999, \>90,000                                                  |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, Other |
|                      |                                                                                                                                 |

@@

For each survey participant certain data was automatically captured, including the nearest municipality (city). This resulted in a total of 578 different municipalities within the clean dataset. Due to our interest in analyzing the differences between Health Regions, we assigned the city of each entry to its correspondent Health Region following a multi-step process. Briefly, we used Local Health Integrated Networks (LHINs) to assign a Health Region to each entry in the survey. LHINs were the geographical divisions for health used by  Ontario before the adoption of the Health Regions; because of the lack of a publicly available list of all municipalities within each Health Region, we used a dataset of long-term care homes and LHINs to match each city to LHIN, followed by matching each LHIN to a Health Region following the information provided on the Ontario Health Website, where the list of LHINs and corresponding Health Regions is available. In the case of municipalities that did not appear in the long-term dataset, we manually searched each city in the LHINs websites in order to provide geographical information. Details of the complete data cleaning process are in the Appendix.

>>>>

| Variable             | Values                                                                                                                          |
|:------------------|-----------------------------------------------------|
| Age group            | 16-24,25-34,35-44,45-54,55-64, 65+                                                                                              |
| Income bracket (CAD) | \<15,000, 15,000-24,999, 25,000-39,999, 40,000-59,999, 60,000-89,999, \>90,000                                                  |
| Race/ethnicity       | Arab/Middle Eastern, Black, East Asian/Pacific Islander, Indigenous, Latin American, Mixed, South Asian, White Caucasian, Other |
|                      |                                                                                                                                 |

@@

For each survey participant certain data was automatically captured, including the nearest municipality (city). This resulted in a total of 578 different municipalities within the clean dataset. Due to our interest in analyzing the differences between Health Regions, we assigned the city of each entry to its correspondent Health Region following a multi-step process. Briefly, we used Local Health Integrated Networks (LHINs) to assign a Health Region to each entry in the survey. LHINs were the geographical divisions for health used by  Ontario before the adoption of the Health Regions; because of the lack of a publicly available list of all municipalities within each Health Region, we used a dataset of long-term care homes and LHINs to match each city to LHIN, followed by matching each LHIN to a Health Region following the information provided on the Ontario Health Website, where the list of LHINs and corresponding Health Regions is available. In the case of municipalities that did not appear in the long-term dataset, we manually searched each city in the LHINs websites in order to provide geographical information. Details of the data cleaning process are in the github repository: https://github.com/aimundo/Fields_COVID-19/.

### 2023-02-07 manuscript/03-results.qmd
created file with details of data cleaning

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic, geographical factors (by Health Region) and their interactions as predictors. The reference groups in each case were set as follows: 16 to 24 years (age group), White Caucasian (Race), Toronto (Health Region), Over CAD 90,000 (Income). Socio-economic factors with statistically significant odds ratios were the age group of 65 years and over (OR=0.44), persons who identified as Arab/Middle Eastern (OR=0.15), Black (OR=0.17), Indigenous (OR=0.27), Latin American (OR=0.24), South Asian (OR=0.4) and Other Race/Ethnicity (a group that included minorities not found elsewhere, OR=0.17), and those that lived in a household with an income bracket between CAD 25,000-39,999 (OR=0.23).

>>>>

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic, geographical factors (by Health Region) and their interactions as predictors. The reference groups in each case were set as follows: 16 to 24 years (age group), White Caucasian (Race), Toronto (Health Region), Over CAD 90,000 (Income). Socio-economic factors with statistically significant odds ratios were the age group of 65 years and over (OR=0.44), persons who identified as Arab/Middle Eastern (OR=0.15), Black (OR=0.17), Indigenous (OR=0.27), Latin American (OR=0.24), South Asian (OR=0.4) and Other Race/Ethnicity (a group those that identified as Southeast Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere, OR=0.17), and those that lived in a household with an income bracket between CAD 25,000-39,999 (OR=0.23).
